---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mutualinf 

<!-- badges: start -->
[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->

An R library to calculate and decompose the Mutual Information Index (M) introduced to social science literature by Theil and Finizza (1971). The M index is a multigroup segregation measure that is highly decomposable that satisfies the Strong Unit Decomposability (SUD) and Strong Group Decomposability (SGD) properties (Frankel and Volij, 2011; Mora and Ruiz-Castillo, 2011). 

The library:

- Allows calculate the total segregation.
- Allows descompose the total segregation index into the between term and the within term.
- Allows calculate and know the exclusive segregating effect that generate the group variables or the unit variables into the total segregation.
- Allows separate the calculations according to one or more characteristics of the system.
- Deliveries the decompositions information in the general form or in detail form. In this last case delivery information about the proportions and the local segregation index for all categorical combinations of variables when there is a least one dimension for the within parameter.
- Is fast in the compute tasks because uses internally the [`data.table`](https://CRAN.R-project.org/package=data.table) package.
- Uses parallelization methods of the [`parallel`](https://stat.ethz.ch/R-manual/R-devel/library/parallel/doc/parallel.pdf) library to help calculate the index and decompositions. This option allows use more than one CPU cores of the computer to efficiently advantage the resources in the calculation. However this option is available only to Mac, Linux, Unix, and BSD systems but is not available to Windows sytems because the package uses internally the function [`mclapply`](https://www.rdocumentation.org/packages/parallel/versions/3.4.1/topics/mclapply) that is based on the bifurcation. In the Windows case, the calculation is carried out sequentially through [`lapply`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/lapply) function.

## Installation

You can install the released version of mutualinf from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("mutualinf")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("RafaelFuentealbaC/mutualinf")
```

## Functions
The package provides two principal functions:

```r
?prepare_data 
```
- This function allows prepare the dataset that later will be use to computation index and generate the decompositions. For more details see the help. 

```r
?mutual
```
- This function provides all values related with computation of the index and decompositions. For more details consult see the help.

## Usage

The package allow calculate the Mutual Information Index (M) in his simplest form, i.e., over one group dimension and one unit dimension:
```{r}
library(mutualinf)

mutual(data = DT_Seg_Chile, group = "csep", unit = "school")
```

Also can be used multiple dimensions in the group and unit analysis:
```{r}
# over multiple group dimensions
mutual(data = DT_Seg_Chile, group = c("csep", "ethnicity"), unit = "school")

# over multiple unit dimensions
mutual(data = DT_Seg_Chile, group = "csep", unit = c("school", "sch_type"))
```

The `by` option allows separate the calculations according to particular dimension or multiple dimensions:
```{r}
mutual(data = DT_Seg_Chile, group = c("csep", "ethnicity"), unit = "school", by = "region")
```

The `within` option allows decompose the total segregation into their between and within terms:
```{r}
# get the segregation that is socio-economic exclusively and then segregation that is ethnic exclusively
# for all socio-economic categories
mutual(data = DT_Seg_Chile, group = c("csep", "ethnicity"), unit = "school", within = "csep")

# get the segregation that is ethnic exclusively and then segregation that is socio-economic exclusively
# for all ethnic categories
mutual(data = DT_Seg_Chile, group = c("csep", "ethnicity"), unit = "school", within = "ethnicity")
```

The `contribution.from` option allows evaluate the exclusive segregating effect of group variables or unit variables into the total segregation. It's an inmediate way of jointly obtaining the relevant results of the two previous decompositions. The `ìnteraction` term refers to an amount of segregation that cannot be attributed to the exclusive segregating effect of characteristics that jointly define the groups (in this case):
```{r}
mutual(data = DT_Seg_Chile, group = c("csep", "ethnicity"), unit = "school", contribution.from = "group_vars")
```

The `components` option allows know the proportions and the local segregation index for all categories of variables of the `within` parameter. The weighted average between `p` and `within` of the `$W_Decomposition` element is equal to the `M_W_csep` term of the `$Total` element:
```{r}
mutual(data = DT_Seg_Chile, group = c("csep", "ethnicity"), unit = "school", within = "csep", components = TRUE)
```

The `cores` option allows use more than one CPU cores in the index compute. This avoids overloading the current work session by distributing calculation tasks in child processs. Compare the differences with the `system.time` function:
```{r}
# Sequentially, using one CPU core:
system.time(mutual(data = DT_Seg_Chile, group = c("csep", "ethnicity", "gender"), unit = c("school", "sch_type", "rural"),
                   within = c("ethnicity", "csep"), contribution.from = "unit_vars", components = TRUE))

# In parallel, using two CPU cores:
system.time(mutual(data = DT_Seg_Chile, group = c("csep", "ethnicity", "gender"), unit = c("school", "sch_type", "rural"),
                   within = c("ethnicity", "csep"), contribution.from = "unit_vars", components = TRUE, cores = 2))

```


## Citation

If you use this library for your research, please cite:

Fuentealba-Chaura, R., Mora, R., Rojas-Mora, J. (2021). mutualinf: a library to calculate and decompose the Mutual Information Index. https://www.github.com/RafaelFuentealbaC/mutualinf


## References

Elbers, B. (2021). A Method for Studying Differences in Segregation Across Time and Space. Sociological Methods & Research. https://doi.org/10.1177/0049124121986204.

Frankel, D. and Volij, O. (2011). Measuring school segregation. Journal of EconomicTheory. 146(1):1-38. https://doi.org/10.1016/j.jet.2010.10.008.

Kullback, S. (1959).Information Theory and Statistics. Wiley Publication in Mathematical Statistics.

Mora, R. and Guinea-Martin, D. (2021). Computing decomposable multigroup indexesof segregation. UC3M Working papers. Economics 31803, Universidad Carlos III de Madrid. Departamento de Economía.

Mora, R. and Ruiz-Castillo, J. (2003). Additively decomposable segregation indexes. The case of gender segregation by occupations and human capital levels in Spain. Journal of Economic Inequality. 1(2):147-179. https://doi.org/10.1023/A:1026198429377.

Mora, R. and Ruiz-Castillo, J. (2011). Entropy-based segregation indices. Sociological Methodology. 41(1):159-194. https://doi.org/10.1111/j.1467-9531.2011.01237.x.

Theil, H. and Finizza, Anthony J. (1971). A note on the measurement of racial integration of schools by means of informational concepts. The Journal of Mathematical Sociology. 1(2):187-193. https://doi.org/10.1080/0022250X.1971.9989795.
